---
title: "AlgorithmComparisonSim"
output: html_document
---

```{r Libraries}

library(tidyverse)
library(scater)
library(splatter)
library(Seurat)
library(cluster)

```

```{r Functions}


# takes parameters specified and returns a number of seurat objects to those specifications, just in a list. they'll be slightly different since theres randomness and we make sure the random seed is different each time. 

make_a_dataset <- function(proj_name, batchCells, group.prob, de.prob, nGenes) {
  
  # batchCells - a vector giving the number of cells per batch, ie two batches with 100 cells is (100, 100)
  # group.prob - vector specifying probabilities per group, ie (0.5, 0.5) will likely give two same sized groups
  # de.prob - probability that genes will be DE by group(?) i.e. that they'll differ more than just noise
  # nGenes - number of genes to simulate
  # proj_name - character name of dataset or for the parameter object
  # all else we leave as default values. 
  
  # a bit of input validation
  stopifnot(
    is.numeric(batchCells),
    #is.numeric(group.prob),
    is.numeric(de.prob),
    is.numeric(nGenes),
    is.character(proj_name)
  ) 
  
  # initialise and set params
  params <- newSplatParams()
  
  params <- setParams(params, update = list(
    batchCells = batchCells, 
    group.prob = group.prob, 
    de.prob = de.prob,
    nGenes = nGenes
    ))
  
  # simulate based on those parameters
  sim <- splatSimulate(params, method = "groups")
  
  counts_matrix <- counts(sim)
  
  # this is what we will write out, and then create seurat object as needed for clustering
  return(counts_matrix)
  
}



```

At present my biggest concern with the below data is that the SCTransform is hanging on to only the 100 most variable genes. I have not found this answer immediate from documentation so I need to dig a bit more for that, or maybe am missing something. 

the columns i want to have in param matrix are:
ID, projname, batchcells, group prob, de prob, nGenes (proj name will be the same as filename)

```{r Make Parameter Matrix}

parameter_matrix <- tibble(
  ID = 1:10,
  proj_name = paste0("set", 1:10),
  batchCells = rep(500, 10),
  group.prob = list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ),
  de.prob = c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5),
  nGenes = rep(2000, 10)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/metadata_acs")) {
  
  cat("metadata_acs file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix, file = "D:/AlgorithmComparisonSim/metadata_acs")
  
  cat("metadata file created. \n")
  
}

```

```{r Generate Datasets}

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  # check if there is a file in the directory by the name we want
  
  if (file.exists(
    
    file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i])
    
  )) {
    
    cat("File: ", metadata$proj_name[i], "already exists in folder. Moving on... \n")
    
  } else {
    
    temp_object <- make_a_dataset(
      proj_name = metadata$proj_name[i],
      batchCells = metadata$batchCells[i],
      group.prob = metadata$group.prob[[i]],
      de.prob = metadata$de.prob[i],
      nGenes = metadata$nGenes[i]
    )
    
    fwrite(temp_object, file = paste0("D:/AlgorithmComparisonSim/", metadata[i, 2]))
    
    cat("File: ", metadata$proj_name[i], "created. Moving on... \n")
    
  }
  
}

```

Since we chose to store the counts instead of the whole seurat object, we will write a quick helper function that brings us to the seurat object. Because of the scaling etc, at this stage we make the decision to cluster all of our methods on the seurat's scaled data layer as far as possible. This will involve some further layers (graphs) for the Louvain and Leiden algorithms but that's fine. 

```{r More Functions}

# rewrite our K Means function. Need to fully understand what's going on. 

run_kmeans <- function(set_index, counts, k = NULL) {
  
  start_time <- Sys.time()
  cat("Started run_kmeans function. \n")
  
  # firstly, we want to take our counts and then make it into a seurat object.
  sobject <- CreateSeuratObject(counts = counts, assay = "RNA", project = paste0("set", set_index))
  
  # then, we want to run SCTransform on the data. To be clear, this does normalisation, scaling, and finding variable features. It creates a whole new assay, which we make sure to reference in further steps. 
  
  sctobject <- SCTransform(sobject)
  
  # and now we look at only the sct scaled  layer from the SCT assay. This is a counts matrix that has been SCTransformed, and it's the scaled layer instead of the raw counts. 
  
  sct_scaled <- LayerData(
    object = sctobject, 
    assay = "SCT", 
    layer = "scale.data")
  
  cat("Object made, transformed... \n")
  
  # and at this stage, we check whether we were passed the true clusters, and if not we estimate it using clusgap. This is a pretty time-intensive step in practice, but we keep the bootstrapping high unless it makes it completely unfeasible. 
  if (is.null(k)) {
    
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = sct_scaled,
      FUNcluster = kmeans,
      K.max = 10,
      B = 20
    )
    
    use_k <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
    
    time_2 <- Sys.time()
    cat("took ", time_2 - time_1, " to estimate k \n")
    
  } else {
    
    use_k <- k
    
  }
  
  # having now estimated or been passed k, we can run kmeans on that same sct_scaled layer. 
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(sct_scaled,
                     centers = use_k,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  # and now having produced our cluster assignments we write them out. For now I think I'll only need the cluster assignments, so we literally just write those out. We will do that in the chunk, and not within this function; here, we return the whole kmobject. 
  
  return(kmobject)
  
}



# run_kmeans <- function(sct_seurat_object, k = NULL) {
#   
#   start_time <- Sys.time()
#   
#   sct_scaled <- LayerData(
#     object = sct_seurat_object, 
#     assay = "SCT", 
#     layer = "scale.data")
# 
#   if (is.numeric(k) == TRUE) {
#     centers = k
#   } else {
#     
#     time_1 <- Sys.time()
#     
#     clusGapcall <- clusGap(
#       x = sct_scaled,
#       FUNcluster = kmeans,
#       K.max = 10,
#       B = 20
#     )
#     
#     centers <- maxSE(clusGapcall$Tab[, "gap"],
#                      clusGapcall$Tab[, "SE.sim"],
#                      method = "firstSEmax")
#     
#     time_2 <- Sys.time()
#     cat("took ", time_2 - time_1, " to estimate k \n")
#     
#   }
#   
#   time_3 <- Sys.time()
#   
#   # run kmeans:
#   kmobject <- kmeans(sct_scaled,
#                      centers = centers,
#                      nstart = 25)
#   
#   end_time <- Sys.time()
#   cat("took ", end_time - time_3, " to cluster and ", end_time - start_time, "in total \n")
#   
#   return(kmobject)
#   
# }

# Takes a seurat object post-SCTransform. Estimates a workable number of PCs, runs PCA, clusters on that number of PCs.

run_PCAkmeans <- function(sct_seurat_object, k = NULL) {
  
  # sct_seurat_object - object of class seurat with SCTransform applied
  # k - centers for KMeans. If left NULL, estimated with Gap Statistic
  
  PCA_sct_seurat_object <- RunPCA(sct_seurat_object, npcs = 100, verbose = TRUE)
  
  # pull PCA embeddings
  embeddings <- Embeddings(PCA_sct_seurat_object)
  
  # run kmeans

  if (is.numeric(k) == TRUE) {
    centers = k
  } else {
    clusGapcall <- clusGap(
      x = embeddings,
      FUNcluster = kmeans,
      K.max = 10,
      B = 5
    )
    
    centers <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
  }
  
  # run kmeans:
  kmobject <- kmeans(embeddings,
                     centers = centers,
                     nstart = 25)
  
  return(kmobject)
  
}

# Takes a seurat object post-SCTransform. Runs the Louvain Algorithm native to Seurat, IF we can make sure it's not doing any extra optimisation under the hood. Otherwise, we can pull the count data, make the SNN graph and cluster ourselves.

run_louvain <- function(sct_seurat_object, resolution  = NULL) {
  
  sct_seurat_object <- sct_seurat_object |> 
    RunPCA(npcs = 100) |> 
    FindNeighbors(dims = 1:30)
    
  
  
  lv_sct <- FindClusters(sct_seurat_object, algorithm = 1)
  
  return(lv_sct)
  
}

```


So far we're seeing the whole thing does run, but the clusgap is returning one centre. am going to run it through all the sets to see if it's always doing this, and then try and tweak it. it will unfortunately probably take 5ish minutes per one, so we'll go with just the first five or so.

Yep so they are all on one centre. So we need to check what's going wrong with our clusgap call, or maybe it's just like too many dimensions. who knows eh.
```{r Giving it a Go}

# set1 <- fread("D:/AlgorithmComparisonSim/set1")
# 
# kmobject <- run_kmeans(1, set1)

t_0 <- Sys.time()

for (i in 1:4) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/set", i))
  
  kmobject <- run_kmeans(i, temp_set)
  
}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 5: ", t_1 - t_0)

```





Not sure about this bit?
```{r Simulate}

kmeans_sct_sim1 <- run_kmeans(sct_sim1)

pca_kmeans_sct_sim1 <- run_PCAkmeans(sct_sim1)

lv_sct_sim1 <- run_louvain(sct_sim1)


# for the louvain clustering

p_sct_sim1 <- sct_sim1 |> 
    RunPCA(npcs = 100) |> 
    FindNeighbors(dims = 1:30)

embeddings <- Embeddings(p_sct_sim1, reduction = "pca")[, 1:30]

lv_clusters <- as.numeric(Idents(sct_sim1))
dist_mat <- dist(embeddings)
lv_sil <- silhouette(lv_clusters, dist_mat)
lv_avg_sil <- mean(lv_sil[, 3])

# then need to run louvain on this




```




