---
title: "AlgorithmComparisonSim"
output: html_document
---

```{r Libraries}

library(tidyverse)
library(scater)
library(splatter)
library(Seurat)
library(cluster)
library(data.table)
library(fpc)
library(clusterSim)

```

```{r Functions}


# takes parameters specified and returns a number of seurat objects to those specifications, just in a list. they'll be slightly different since theres randomness and we make sure the random seed is different each time. 

make_a_dataset <- function(proj_name, batchCells, group.prob, de.prob, nGenes) {
  
  # batchCells - a vector giving the number of cells per batch, ie two batches with 100 cells is (100, 100)
  # group.prob - vector specifying probabilities per group, ie (0.5, 0.5) will likely give two same sized groups
  # de.prob - probability that genes will be DE by group(?) i.e. that they'll differ more than just noise
  # nGenes - number of genes to simulate
  # proj_name - character name of dataset or for the parameter object
  # all else we leave as default values. 
  
  # a bit of input validation
  stopifnot(
    is.numeric(batchCells),
    #is.numeric(group.prob),
    is.numeric(de.prob),
    is.numeric(nGenes),
    is.character(proj_name)
  ) 
  
  # initialise and set params
  params <- newSplatParams()
  
  params <- setParams(params, update = list(
    batchCells = batchCells, 
    group.prob = group.prob, 
    de.prob = de.prob,
    nGenes = nGenes
    ))
  
  # simulate based on those parameters
  sim <- splatSimulate(params, method = "groups")
  
  counts_matrix <- counts(sim)
  
  # extract true labelling for later
  true_labels <- data.table(
    cell_id = colnames(counts_matrix),
    true_group = colData(sim)$Group
  )
  
  # this is what we will write out, and then create seurat object as needed for clustering from the counts, and will use true labels for the arandi later.
  return(list(
    counts = counts_matrix,
    labels = true_labels
  ))
  
}

#############################################################################

# rewrite our K Means function. Need to fully understand what's going on. 

run_kmeans <- function(set_index, counts, k = NULL) {
  
  start_time <- Sys.time()
  cat("BEGIN run_kmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes
  counts <- t(counts)
  
  # we remove dimensions that dont vary at all, as they won't help with clustering
  counts <- counts[, apply(counts, 2, var) > 0]
  
  # I think that we want to scale the data in order for us to see the clusters
  # counts <- scale(counts)
  
  # Estimate k with clusGap
  if (is.null(k)) {
    
    cat("Estimating K... \n")
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = counts,
      FUNcluster = kmeans,
      K.max = 10,
      B = 5, #REDUCED TEMPORARILY, RESET TO 20
      nstart = 5 #REDUCED TEMPORARILY< RESET TO 25
    )
    
    use_k <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
    
    time_2 <- Sys.time()
    cat("took ", time_2 - time_1, " to estimate K as ", use_k, " \n")
    
  } else {
    
    use_k <- k
    cat("K passed as", use_k)
    
  }
  
  # having now estimated or been passed k, we can run kmeans on that same sct_scaled layer. 
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(counts,
                     centers = use_k,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  # and now having produced our cluster assignments we write them out. For now I think I'll only need the cluster assignments, so we literally just write those out. 
  return(kmobject$cluster)
  
  cat("END run_kmeans function. \n")
  
}

#############################################################################


# Takes a matrix of counts data (transposes it) and then finds a good number of PCs by which to split the data for cluster evaluation. Returns a list of this number and a plot of the variance explained by each PC, so that we can easily compare whether it's doing the right thing. 
choose_pcs_elbow <- function(dataset) {
  
  counts <- t(dataset)
  prcobject <- prcomp(counts)
  variances <- prcobject$sdev^2
  n_pcs <- length(variances)
  
  x <- 1:n_pcs
  y <- variances
  
  x1 <- x[1]
  y1 <- y[1]
  x2 <- x[n_pcs]
  y2 <- y[n_pcs]
  
  a <- y2 - y1
  b <- -(x2 - x1)
  c <- (x2 - x1) * y1 - (y2 - y1) * x1
  
  distances <- abs(a * x + b * y + c) / sqrt(a^2 + b^2)
  
  # The elbow is at the point with maximum distance
  # this is a pretty heavy assumption
  elbow_idx <- which.max(distances)
  
  elbowplot <- ggplot(mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_line() +  # Optional: connect the points
  geom_abline(intercept = y[1], 
              slope = (y[2] - y[1]) / n_pcs,
              linetype = "dashed", 
              color = "gray50") +
  theme_minimal()
  
  return(list(elbow_idx = elbow_idx,
              elbowplot = elbowplot))
  
}

#############################################################################

choose_pcs_threshold <- function(dataset) {
  
  # transpose then drop zero variance columns from consideration
  dataset <- t(dataset)
  dataset <- dataset[, apply(dataset, 2, var) != 0]
  
  # Calculate PCA
  pca_result <- prcomp(dataset, center = TRUE, scale. = TRUE)
  
  # Create data frame for plotting
  var_explained <- data.frame(
    PC = 1:length(pca_result$sdev),
    Variance = pca_result$sdev^2 / sum(pca_result$sdev^2),
    Cumulative = cumsum(pca_result$sdev^2 / sum(pca_result$sdev^2))
  )
  
  # Plot
  plot <- ggplot(var_explained, aes(x = PC, y = Variance)) +
    geom_line() +
    geom_point(size = 2) +
    labs(x = "Principal Component",
         y = "Proportion of Variance Explained",
         title = "PCA Variance Explained") +
    theme_minimal()
  
  first_past_threshold <- var_explained$PC[which(var_explained$Cumulative > 0.8)[1]]
  
  return(list(var_explained = var_explained,
              plot = plot,
              first_past_threshold = first_past_threshold))
  
}

#############################################################################


estimate_k <- function(embeddings, scale) {
  
  if (!is.logical(scale) | scale == FALSE) {
    cat("Not scaling embeddings... \n")
  } else if (scale == TRUE) {
    embeddings <- scale(embeddings)
    cat("Scaling embeddings... \n")
  }
    
  clusGapcall <- clusGap(embeddings,
                       FUNcluster = kmeans,
                       nstart = 25,
                       iter.max = 20,
                       K.max = 10,
                       B = 100)

  k_val <- maxSE(clusGapcall$Tab[, "gap"], clusGapcall$Tab[, "SE.sim"], 
             method = "globalmax")
  
  # Specify 2 as a floor for clustering such that we can extract a silhouette score from it. This is hopefully wrapped up in our assumption that there are clusters at all, so it makes sense to specify there are more than one. 
  if (k_val < 2) {
    k_val <- 2
    cat("Had to use the 2 cluster floor btw... \n")
  }

  cat("clusGap gives ", k_val, " as our K estimate. \n")
  
  return(k_val)
  
}



##############################################################################


# Takes a seurat object post-SCTransform. Estimates a workable number of PCs, runs PCA, clusters on that number of PCs.

run_PCAkmeans <- function(set_index, counts, threshold, k = NULL, scale = FALSE) {
  
  start_time <- Sys.time()
  cat("BEGIN run_PCAkmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes and get rid of the invariant cols
  counts <- t(counts)
  counts <- counts[, apply(counts, 2, var) > 0]
  
  npcs <- threshold
  
  cat("Continuing on ", npcs, " PCs. \n")
  
  embeddings <- prcomp(counts)$x[, 1:npcs]

  if (is.null(k)) {
    
    time_1 <- Sys.time()
    
    use_k <- estimate_k(embeddings = embeddings, scale)
    
    time_2 <- Sys.time()
    cat("Took ", time_2 - time_1, " to estimate k. \n")
    
  } else {
    
    use_k <- k
    cat("K passed as ", use_k, " \n")
    
  }
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(embeddings,
                     centers = use_k,
                     iter.max = 20,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  #r returning just the cluster vector for now
  
  clustering <- data.frame(
    cell = names(kmobject$cluster),
    cluster = kmobject$cluster,
    row.names = NULL
  )
  
  return(clustering)
  
  cat("END run_PCAkmeans function. \n")
  
}


##############################################################################


# Takes a seurat object post-SCTransform. Runs the Louvain Algorithm native to Seurat, IF we can make sure it's not doing any extra optimisation under the hood. Otherwise, we can pull the count data, make the SNN graph and cluster ourselves.

run_graph_based <- function(set_index, counts, mode, npcs) {
  # set_index - the index of the dataset to be used
  # counts - the counts dataset to be used
  # mode - 1 for Louvain, 2 for Leiden
  
  if (mode == "Louvain") {
    # choose Louvain
    algorithm <- 1
  } else if (mode == "Leiden") {
    # choose Leiden
    algorithm <- 4
  } else {
    algorithm <- 1
    cat("Mode not specified correctly - using Louvain. \n")
  }
  
  start_time <- Sys.time()
  cat("BEGIN run_graph_based function. \n")
  
  # firstly, we want to take our counts and then make it into a seurat object.
  sobject <- CreateSeuratObject(counts = counts, assay = "RNA", project = paste0("set", set_index))
  
  # then, we want to run SCTransform on the data. To be clear, this does normalisation, scaling, and finding variable features. It creates a whole new assay, which we make sure to reference in further steps. 
  
  sctobject <- SCTransform(sobject)
  
  sct_scaled <- LayerData(object = sctobject, assay = "SCT", layer = "scale.data")

  cat("Continuing", mode, " on ", npcs, " PCs... \n")
  
  sctobject <- sctobject |> 
    RunPCA(npcs = npcs) |> 
    FindNeighbors(dims = 1:npcs)
    
  # select Louvain with Algorithm = 1
  sctobject <- FindClusters(sctobject, algorithm = algorithm)
  
  clustering <- data.frame(cell = names(sctobject$seurat_clusters), cluster = as.character(sctobject$seurat_clusters))
  
  cat("END run_graph_based function. \n")
  
  return(clustering)
  
}

###############################################################################

cluster_scores <- function(clustering, original_counts, true_labels, algorithm) {
  
  # silhouette score
  whole_silhouette <- silhouette(clustering, dist = dist(original_counts))
  score_silhouette <- mean(whole_silhouette[, 3])
  
  #calinhara 
  score_calinhara <- calinhara(original_counts, clustering)
  
  #dbouldin
  score_dbouldin <- index.DB(original_counts, clustering)[["DB"]]
  
  #arandi
  score_ari <- mclust::adjustedRandIndex(true_labels, clustering)
  
  scores <- tibble(
    
    algorithm = algorithm,
    avg_silhouette = score_silhouette,
    calinski_harabasz = score_calinhara,
    davies_bouldin = score_dbouldin,
    adjusted_rand_index = score_ari
    
  )
  
  return(scores)
  
}

```

reference to github can go at end of intro or similar

%%%%%%%%%%%%%%%%%%%
% Base Simulation %
%%%%%%%%%%%%%%%%%%%

```{r Make Parameter Matrix}

parameter_matrix <- tibble(
  ID = 1:60,
  proj_name = paste0("set", sprintf("%02d", 1:60)),
  batchCells = rep(500, 60),
  group.prob = rep(list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ), times = 6),
  de.prob = rep(c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5), each = 6),
  nGenes = rep(2000, 60)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/BaseSim/metadata_acs")) {
  
  cat("metadata_acs file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix, file = "D:/AlgorithmComparisonSim/BaseSim/metadata_acs")
  
  cat("metadata file created. \n")
  
}

```

```{r Generate Datasets}

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  # check if there is a file in the directory by the name we want
  
  if (!file.exists(file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i])) | 
      !file.exists(file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i], "true_labels"))) {
    
    # make the dataset if at least one of them is missing
    temp_object <- make_a_dataset(
      proj_name = metadata$proj_name[i],
      batchCells = metadata$batchCells[i],
      group.prob = metadata$group.prob[[i]],
      de.prob = metadata$de.prob[i],
      nGenes = metadata$nGenes[i]
    )
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i]))) {
      cat("File for", i, "already exists, moving on... \n")
    } else {
      
      #write out counts
      fwrite(temp_object$counts, file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata[i, 2]))
      
      cat("File: ", metadata$proj_name[i], "created. Moving on... \n")
      
    }
    
    
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i], "true_labels"))) {
      cat("File: ", metadata$proj_name[i], "true_labels already exists in folder. Moving on... \n")
      
      } else {
        
      
        fwrite(temp_object$labels, file = paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i], "true_labels"))  
        
        cat("True labels for", i, "created... \n")
        
    }
    
  }
  
}

```

```{r Choose Number of PCs}

# read the file back in and set random seed for reproducibility
metadata_acs <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata_acs <- metadata_acs |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

# initialise vector
threshold_values <- matrix(nrow = nrow(metadata_acs), ncol = 1)

for (i in 1:nrow(metadata_acs)) {
  
  temp_set <- fread(paste0("D:AlgorithmComparisonSim/BaseSim/", metadata_acs$proj_name[[i]]))
  
  value <- choose_pcs_elbow(temp_set)$elbow_idx

  #cat(value, " \n")
  threshold_values[i] <- value
  cat(".")
}

mean(threshold_values)

```
We go with 10, as with our elbow method we can justify it and it seems to be giving the correct cluster numbers. Great!

```{r KMeans Cluster and Writeout}

t_0 <- Sys.time()

cat("BEGINNING PCA KMEANS CLUSTERING FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/BaseSim/", metadata$proj_name[i]
  ))
  
  
  
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/BaseSim/PK", sprintf("%02d", i))
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/BaseSim/PK", sprintf("%02d", i)), "already exists... \n") 
  } else {
    time_a <- Sys.time()
    # We scale the PC embeddings because it magically works, yay!
    cat("Clustering the", i, "th set... \n")
    temp_cluster <- run_PCAkmeans(i, temp_set, threshold = 10, scale = TRUE)
    cat("Writing out the", i, "th set... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/BaseSim/PK", sprintf("%02d", i)))
    time_b <- Sys.time()
    cat("Written out file from set", i, " and took ", time_b - time_a, "to do that set in total ... \n")
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all em. \n")

```
Done!

```{r Louvain Cluster and Writeout}

t_0 <- Sys.time()

cat("BEGINNING LOUVAIN CLUSTERING FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acs)) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata_acs$proj_name[i]))
  
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/BaseSim/LV", sprintf("%02d", i))
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/BaseSim/LV", sprintf("%02d", i)), "already exists... \n") 
  } else {
    time_a <- Sys.time()
    cat("Clustering... \n")
    temp_cluster <- run_graph_based(i, temp_set, mode = "Louvain", npcs = 10)
    cat("Writing.. \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/BaseSim/LV", sprintf("%02d", i)))
    time_b <- Sys.time()
    cat("Written out file from set", i, "and took", time_b - time_a, "to do that... \n")
    }

}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 10: ", t_1 - t_0)

```
Done!

```{r Leiden Cluster and Writeout}

t_0 <- Sys.time()

cat("BEGINNING LEIDEN CLUSTERING FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acs)) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/", metadata_acs$proj_name[i]))
  
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/BaseSim/LD", sprintf("%02d", i))
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/BaseSim/LD", sprintf("%02d", i)), "already exists... \n") 
  } else {
    time_a <- Sys.time()
    cat("Clustering... \n")
    temp_cluster <- run_graph_based(i, temp_set, mode = "Leiden", npcs = 10)
    cat("Writing.. \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/BaseSim/LD", sprintf("%02d", i)))
    time_b <- Sys.time()
    cat("Written out file from set", i, "and took", time_b - time_a, "to do that... \n")
    }

}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 10: ", t_1 - t_0)

```
Done!

```{r PK Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING PK SCORES WRITEOUT... \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

PK_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/PK", sprintf("%02d", i)))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "PCAKMeans",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  PK_allscores <- add_row(PK_allscores, temp_score)
  cat("added", i, "one to PK_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/BaseSim/PK_allscores")) {
  cat("Scores for PK already exist... \n")
} else {
  fwrite(PK_allscores, "D:/AlgorithmComparisonSim/BaseSim/PK_allscores")
  cat("Written out PK scores... \n")
}

```
Done!

```{r LV Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING LV SCORES WRITEOUT... \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LV_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/LV", sprintf("%02d", i)))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Louvain",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LV_allscores <- add_row(LV_allscores, temp_score)
  cat("added", i, "one to LV_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/BaseSim/LV_allscores")) {
  cat("Scores for LV already exist... \n")
} else {
  fwrite(LV_allscores, "D:/AlgorithmComparisonSim/BaseSim/LV_allscores")
  cat("Written out LV scores... \n")
}



```
Done!

```{r LD Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING LD SCORES WRITEOUT... \n")

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/BaseSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LD_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/LD", sprintf("%02d", i)))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/BaseSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Leiden",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LD_allscores <- add_row(LD_allscores, temp_score)
  cat("added", i, "one to LD_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/BaseSim/LD_allscores")) {
  cat("Scores for LD already exist... \n")
} else {
  fwrite(LD_allscores, "D:/AlgorithmComparisonSim/BaseSim/LD_allscores")
  cat("Written out LD scores... \n")
}

```
Done!

```{r Comparison Plot}

LV_allscores <- fread("D:/AlgorithmComparisonSim/BaseSim/LV_allscores")
LD_allscores <- fread("D:/AlgorithmComparisonSim/BaseSim/LD_allscores")
PK_allscores <- fread("D:/AlgorithmComparisonSim/BaseSim/PK_allscores")


combined_scores <- bind_rows(LV_allscores, LD_allscores, PK_allscores)


# Reshape data from wide to long format
combined_scores_long <- combined_scores %>%
  pivot_longer(
    cols = c(avg_silhouette, calinski_harabasz, davies_bouldin, adjusted_rand_index),
    names_to = "metric",
    values_to = "value"
  )

# Create faceted violin plots
plot1 <- ggplot(combined_scores_long, aes(x = algorithm, y = value, fill = algorithm)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +  
  theme_bw() +
  labs(
    title = "Clustering Performance Metrics by Algorithm",
    x = "Algorithm",
    y = "Score"
  ) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_brewer(palette = "Set1")  

plot1 

# ggsave("C:/Users/fmhfl/Desktop/Undergrad Project/undergrad_project/OUTPUTS/metrics_by_algorithm.png", 
#        plot = plot1, 
#        width = 10, 
#        height = 8, 
#        dpi = 300)

```
Done!

%%%%%%%%%%%%%%%%%%%%%%%%%
% Separation Simulation %
%%%%%%%%%%%%%%%%%%%%%%%%%

```{r Make Parameter Matrix SEP}

parameter_matrix_separation <- tibble(
  ID = 1:60,
  proj_name = paste0("set", sprintf("%02d", 1:60), "SEP"),
  batchCells = rep(500, 60),
  group.prob = rep(list(c(0.25, 0.25, 0.25, 0.25)), 60),
  de.prob = c(
    rep(0.1, 15),
    rep(0.2, 15),
    rep(0.3, 15),
    rep(0.4, 15)
  ),
  nGenes = rep(2000, 60)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")) {
  
  cat("metadata_acsSEP file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix_separation, file = "D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")
  
  cat("metadata_acsSEP file created. \n")
  
}

```
Done!

```{r Generate Datasets SEP}

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsSEP)) {
  
  # check if there is a file in the directory by the name we want
  
  if (!file.exists(file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i])) | 
      !file.exists(file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i], "true_labels"))) {
    
    # make the dataset if at least one of them is missing
    temp_object <- make_a_dataset(
      proj_name = metadata_acsSEP$proj_name[i],
      batchCells = metadata_acsSEP$batchCells[i],
      group.prob = metadata_acsSEP$group.prob[[i]],
      de.prob = metadata_acsSEP$de.prob[i],
      nGenes = metadata_acsSEP$nGenes[i]
    )
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i]))) {
      cat("File for", i, "already exists, moving on... \n")
    } else {
      
      #write out counts
      fwrite(temp_object$counts, file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP[i, 2]))
      
      cat("File: ", metadata_acsSEP$proj_name[i], "created. Moving on... \n")
      
    }
    
    
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i], "true_labels"))) {
      cat("File: ", metadata_acsSEP$proj_name[i], "true_labels already exists in folder. Moving on... \n")
      
      } else {
        
      
        fwrite(temp_object$labels, file = paste0("D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i], "true_labels"))  
        
        cat("True labels for", i, "created... \n")
        
    }
    
  }
  
}


```
Done!

```{r Choosing a Number of PCs SEP}

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

# initialise vector
threshold_values <- matrix(nrow = nrow(metadata_acsSEP), ncol = 1)

for (i in 1:nrow(metadata_acsSEP)) {
  
  temp_set <- fread(paste0("D:AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[[i]]))
  
  value <- choose_pcs_elbow(temp_set)$elbow_idx

  #cat(value, " \n")
  threshold_values[i] <- value
  cat(".")
}

mean(threshold_values)


```
Having chosen via an elbow method, we obtain an average of just above 7. Hence we continue on 10.

```{r PK SEP Cluster and Writeout}

t_0 <- Sys.time()
cat("BEGINNING PK SEP CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsSEP)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/SeparationSim/PK", sprintf("%02d", i), "SEP")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/SeparationSim/PK", sprintf("%02d", i), "SEP"), "already exists... \n") 
  } else {
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    # DO scale our embeddings
    temp_cluster <- run_PCAkmeans(i, temp_set, threshold = 10, scale = TRUE)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/SeparationSim/PK", sprintf("%02d", i), "SEP"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
  
  cat("FINISHED with the ", i, "th one! \n")
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 60. \n")

```
Done!

```{r LV SEP Cluster and Writeout}

t_0 <- Sys.time()
cat("BEGINNING LV SEP CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsSEP)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/SeparationSim/LV", sprintf("%02d", i), "SEP")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/SeparationSim/LV", sprintf("%02d", i), "SEP"), "already exists... \n") 
  } else {
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    temp_cluster <- run_graph_based(i, temp_set, mode = "Louvain", npcs = 10)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/SeparationSim/LV", sprintf("%02d", i), "SEP"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 10. \n")

```
Done!

```{r LD SEP Cluster and Writeout}

t_0 <- Sys.time()
cat("BEGINNING LD SEP CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsSEP)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/SeparationSim/", metadata_acsSEP$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/SeparationSim/LD", sprintf("%02d", i), "SEP")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/SeparationSim/LD", sprintf("%02d", i), "SEP"), "already exists... \n") 
  } else {
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    # DO scale our embeddings
    temp_cluster <- run_graph_based(i, temp_set, mode = "Leiden", npcs = 10)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/SeparationSim/LD", sprintf("%02d", i), "SEP"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 10. \n")

```
Done!

```{r PK SEP Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING PK SEP SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

PK_SEP_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsSEP)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/PK", sprintf("%02d", i), "SEP"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEP")))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEPtrue_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "PCAKMeans",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  PK_SEP_allscores <- add_row(PK_SEP_allscores, temp_score)
  cat("added", i, "one to PK_SEP_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/SeparationSim/PK_SEP_allscores")) {
  cat("Scores for PK SEP already exist... \n")
} else {
  fwrite(PK_SEP_allscores, "D:/AlgorithmComparisonSim/SeparationSim/PK_SEP_allscores")
  cat("Written out PK SEP scores... \n")
}

```
Done!

```{r LV SEP Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING LV SEP SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LV_SEP_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsSEP)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/LV", sprintf("%02d", i), "SEP"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEP")))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEPtrue_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Louvain",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LV_SEP_allscores <- add_row(LV_SEP_allscores, temp_score)
  cat("added", i, "one to LV_SEP_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/SeparationSim/LV_SEP_allscores")) {
  cat("Scores for LV SEP already exist... \n")
} else {
  fwrite(LV_SEP_allscores, "D:/AlgorithmComparisonSim/SeparationSim/LV_SEP_allscores")
  cat("Written out LV SEP scores... \n")
}

```
Done!

```{r LD SEP Scores Writeout}

t_0 <- Sys.time()
cat("BEGINNING LD SEP SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LD_SEP_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsSEP)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/LD", sprintf("%02d", i), "SEP"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEP")))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/SeparationSim/set", sprintf("%02d", i), "SEPtrue_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Leiden",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LD_SEP_allscores <- add_row(LD_SEP_allscores, temp_score)
  cat("added", i, "one to LD_SEP_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/SeparationSim/LD_SEP_allscores")) {
  cat("Scores for LD SEP already exist... \n")
} else {
  fwrite(LD_SEP_allscores, "D:/AlgorithmComparisonSim/SeparationSim/LD_SEP_allscores")
  cat("Written out LD SEP scores... \n")
}

```
Done!

```{r Comparison Plot SEP}

LV_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/LV_SEP_allscores")
LD_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/LD_SEP_allscores")
PK_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/PK_SEP_allscores")


combined_scores <- bind_rows(LV_SEP_allscores, LD_SEP_allscores, PK_SEP_allscores)

# Reshape data from wide to long format
combined_scores_long <- combined_scores %>%
  pivot_longer(
    cols = c(avg_silhouette, calinski_harabasz, davies_bouldin, adjusted_rand_index),
    names_to = "metric",
    values_to = "value"
  )

# Create faceted violin plots
plot1 <- ggplot(combined_scores_long, aes(x = algorithm, y = value, fill = algorithm)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +  
  theme_bw() +
  labs(
    title = "Clustering Performance Metrics by Algorithm: Separation Simulation",
    x = "Algorithm",
    y = "Score"
  ) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_brewer(palette = "Set1")  

plot1 

# ggsave("C:/Users/fmhfl/Desktop/Undergrad Project/undergrad_project/OUTPUTS/metrics_by_algorithm.png", 
#        plot = plot1, 
#        width = 10, 
#        height = 8, 
#        dpi = 300)

```

```{r Separation Plot SEP}

metadata_acsSEP <- fread("D:/AlgorithmComparisonSim/SeparationSim/metadata_acsSEP")

metadata_acsSEP <- metadata_acsSEP |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LV_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/LV_SEP_allscores")
LV_SEP_allscores$separation <- metadata_acsSEP$de.prob

LD_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/LD_SEP_allscores")
LD_SEP_allscores$separation <- metadata_acsSEP$de.prob

PK_SEP_allscores <- fread("D:/AlgorithmComparisonSim/SeparationSim/PK_SEP_allscores")
PK_SEP_allscores$separation <- metadata_acsSEP$de.prob


combined_scores <- bind_rows(LV_SEP_allscores, LD_SEP_allscores, PK_SEP_allscores)

# Reshape data from wide to long format
combined_scores_long <- combined_scores %>%
  pivot_longer(
    cols = c(avg_silhouette, calinski_harabasz, davies_bouldin, adjusted_rand_index),
    names_to = "metric",
    values_to = "value"
  )

# Create faceted violin plots
plot1 <- ggplot(combined_scores_long, aes(x = algorithm, y = value)) +
  geom_boxplot() +
  geom_jitter(aes(colour = as.factor(separation)), width = 0.1, alpha = 0.5, size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +  
  theme_bw() +
  labs(
    title = "Clustering Performance Metrics by Algorithm: Separation Simulation",
    x = "Algorithm",
    y = "Score"
  ) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_brewer(palette = "Set1")  

plot1 

```

%%%%%%%%%%%%%%%%%%%%%%%%
% PC Number Simulation %
%%%%%%%%%%%%%%%%%%%%%%%%

For this, the data is generated much the same (identically?) to the base sim. However, after finding an optimal number of PCs for the data, we then run the clustering and scoring on 1, .5, and 2 times this number of PCs. The results will show the robustness of different scoring methods to number of PCs. Plan is sixty datasets. 

```{r Make Parameter Matrix PCN}

# similar to base simulation, but want each third to be identical
#each third already is? so our metadata file is just identical to the base sim one. not a problem i don't think. 
parameter_matrix_PCN <- tibble(
  ID = 1:60,
  proj_name = paste0("set", sprintf("%02d", 1:60)),
  batchCells = rep(500, 60),
  group.prob = rep(list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ), times = 6),
  de.prob = rep(c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5), each = 6),
  nGenes = rep(2000, 60)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")) {
  
  cat("metadata_acsPCN file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix_PCN, file = "D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")
  
  cat("metadata_acsPCN file created. \n")
  
}

```

```{r Generate Datasets PCN}

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsPCN)) {
  
  # check if there is a file in the directory by the name we want
  
  if (!file.exists(file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i])) | 
      !file.exists(file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i], "true_labels"))) {
    
    # make the dataset if at least one of them is missing
    temp_object <- make_a_dataset(
      proj_name = metadata_acsPCN$proj_name[i],
      batchCells = metadata_acsPCN$batchCells[i],
      group.prob = metadata_acsPCN$group.prob[[i]],
      de.prob = metadata_acsPCN$de.prob[i],
      nGenes = metadata_acsPCN$nGenes[i]
    )
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i]))) {
      cat("File for", i, "already exists, moving on... \n")
    } else {
      
      #write out counts
      fwrite(temp_object$counts, file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN[i, 2]))
      
      cat("File: ", metadata_acsPCN$proj_name[i], "created. Moving on... \n")
      
    }
    
    
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i], "true_labels"))) {
      cat("File: ", metadata_acsPCN$proj_name[i], "true_labels already exists in folder. Moving on... \n")
      
      } else {
        
      
        fwrite(temp_object$labels, file = paste0("D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i], "true_labels"))  
        
        cat("True labels for", i, "created... \n")
        
    }
    
  }
  
}

```

```{r Choose PC Thresholds PCN}

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

# initialise vector
threshold_values <- matrix(nrow = nrow(metadata_acsPCN), ncol = 1)

for (i in 1:nrow(metadata_acsPCN)) {
  
  temp_set <- fread(paste0("D:AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[[i]]))
  
  value <- choose_pcs_elbow(temp_set)$elbow_idx

  #cat(value, " \n")
  threshold_values[i] <- value
  cat(".")
}

mean(threshold_values)

```
We see the exact same result from the BASE SIM. So we have our three levels: 5, 10, 20 PCs. 


```{r PK Cluster and Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING PK PCN CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsPCN)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/PCNSim/PK", sprintf("%02d", i), "PCN")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/PCNSim/PK", sprintf("%02d", i), "PCN"), "already exists... \n") 
  } else {
    
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    
    # split by PCN
    if (i <= 20) {
      level <- 5
    } else if (i <= 40) {
      level <- 10
    } else {
      level <- 20
    }
    
    temp_cluster <- run_PCAkmeans(i, temp_set, threshold = level, scale = TRUE)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/PCNSim/PK", sprintf("%02d", i), "PCN"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 60. \n")

```
Done!

```{r LV Cluster and Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING LV PCN CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsPCN)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/PCNSim/LV", sprintf("%02d", i), "PCN")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/PCNSim/LV", sprintf("%02d", i), "PCN"), "already exists... \n") 
  } else {
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    
    # split by PCN
    if (i <= 20) {
      level <- 5
    } else if (i <= 40) {
      level <- 10
    } else {
      level <- 20
    }
    
    temp_cluster <- run_graph_based(i, temp_set, mode = "Louvain", npcs = level)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/PCNSim/LV", sprintf("%02d", i), "PCN"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 10. \n")

```
Done!

```{r LD Cluster and Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING LD PCN CLUSTER and WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata_acsPCN)) {
  cat("Reading in the", i, "th dataset... \n")
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/PCNSim/", metadata_acsPCN$proj_name[i]
  ))
  if (file.exists(
    paste0("D:/AlgorithmComparisonSim/PCNSim/LD", sprintf("%02d", i), "PCN")
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/PCNSim/LD", sprintf("%02d", i), "PCN"), "already exists... \n") 
  } else {
    t_a <- Sys.time()
    cat("Clustering begins... \n")
    
    # split by PCN
    if (i <= 20) {
      level <- 5
    } else if (i <= 40) {
      level <- 10
    } else {
      level <- 20
    }
    
    temp_cluster <- run_graph_based(i, temp_set, mode = "Leiden", npcs = level)
    cat("Clustering done, writing out... \n")
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/PCNSim/LD", sprintf("%02d", i), "PCN"))
    cat("Written out file from set", i, " ... \n")
    t_b <- Sys.time()
    cat("Took", t_b - t_a, "for the ", i, "th one... \n")
    
    }
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do all 10. \n")

```
Done!

```{r PK Scores Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING PK PCN SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

PK_PCN_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsPCN)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/PK", sprintf("%02d", i), "PCN"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "PCAKMeans",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  PK_PCN_allscores <- add_row(PK_PCN_allscores, temp_score)
  cat("added", i, "one to PK_PCN_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/PCNSim/PK_PCN_allscores")) {
  cat("Scores for PK PCN already exist... \n")
} else {
  fwrite(PK_SEP_allscores, "D:/AlgorithmComparisonSim/PCNSim/PK_PCN_allscores")
  cat("Written out PK PCN scores... \n")
}

```
Done

```{r LV Scores Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING LV PCN SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LV_PCN_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsPCN)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/LV", sprintf("%02d", i), "PCN"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Louvain",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LV_PCN_allscores <- add_row(LV_PCN_allscores, temp_score)
  cat("added", i, "one to LV_PCN_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/PCNSim/LV_PCN_allscores")) {
  cat("Scores for LV PCN already exist... \n")
} else {
  fwrite(LV_SEP_allscores, "D:/AlgorithmComparisonSim/PCNSim/LV_PCN_allscores")
  cat("Written out LV PCN scores... \n")
}

```
Done

```{r LD Scores Writeout PCN}

t_0 <- Sys.time()
cat("BEGINNING LD PCN SCORES WRITEOUT FULL RUN. \n")

# read the file back in and set random seed for reproducibility
metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LD_PCN_allscores <- tibble(
  algorithm = character(),
  avg_silhouette = numeric(),
  calinski_harabasz = numeric(),
  davies_bouldin = numeric(),
  adjusted_rand_index = numeric()
)

for (i in 1:nrow(metadata_acsPCN)) {
  
  temp_cluster <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/LD", sprintf("%02d", i), "PCN"))[["cluster"]]
  temp_set <- t(fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i))))
  temp_true <- fread(paste0("D:/AlgorithmComparisonSim/PCNSim/set", sprintf("%02d", i), "true_labels"))[["true_group"]]
  
  t_a <- Sys.time()
  cat("Scoring the ", i, "th one... \n")
  
  temp_score <- cluster_scores(algorithm = "Leiden",
    clustering = temp_cluster,
    original_counts = temp_set,
    true_labels = temp_true)
  
  LD_PCN_allscores <- add_row(LD_PCN_allscores, temp_score)
  cat("added", i, "one to LD_PCN_allscores... \n")
  t_b <- Sys.time()
  cat(t_b - t_a, "to do the ", i, "loop... \n")
  
}

t_1 <- Sys.time()
cat("TOOK", t_1 - t_0, "TO DO ALL 10... \n")

if (file.exists("D:/AlgorithmComparisonSim/PCNSim/LD_PCN_allscores")) {
  cat("Scores for LD PCN already exist... \n")
} else {
  fwrite(LD_SEP_allscores, "D:/AlgorithmComparisonSim/PCNSim/LD_PCN_allscores")
  cat("Written out LD PCN scores... \n")
}

```
Done

```{r Comparison Plot PCN}

metadata_acsPCN <- fread("D:/AlgorithmComparisonSim/PCNSim/metadata_acsPCN")

metadata_acsPCN <- metadata_acsPCN |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

LV_PCN_allscores <- fread("D:/AlgorithmComparisonSim/PCNSim/LV_PCN_allscores")
LD_PCN_allscores <- fread("D:/AlgorithmComparisonSim/PCNSim/LD_PCN_allscores")
PK_PCN_allscores <- fread("D:/AlgorithmComparisonSim/PCNSim/PK_PCN_allscores")


combined_scores <- bind_rows(LV_PCN_allscores, LD_PCN_allscores, PK_PCN_allscores)

# Reshape data from wide to long format
combined_scores_long <- combined_scores %>%
  pivot_longer(
    cols = c(avg_silhouette, calinski_harabasz, davies_bouldin, adjusted_rand_index),
    names_to = "metric",
    values_to = "value"
  )

# Create faceted violin plots
plot1 <- ggplot(combined_scores_long, aes(x = algorithm, y = value, fill = algorithm)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +  
  theme_bw() +
  labs(
    title = "Clustering Performance Metrics by Algorithm: Principal Component Number Simulation",
    x = "Algorithm",
    y = "Score"
  ) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_brewer(palette = "Set1")  

plot1 

```

