---
title: "AlgorithmComparisonSim"
output: html_document
---

```{r Libraries}

library(tidyverse)
library(scater)
library(splatter)
library(Seurat)
library(cluster)
library(data.table)
library(fpc)
library(clusterSim)

```

```{r Functions}


# takes parameters specified and returns a number of seurat objects to those specifications, just in a list. they'll be slightly different since theres randomness and we make sure the random seed is different each time. 

make_a_dataset <- function(proj_name, batchCells, group.prob, de.prob, nGenes) {
  
  # batchCells - a vector giving the number of cells per batch, ie two batches with 100 cells is (100, 100)
  # group.prob - vector specifying probabilities per group, ie (0.5, 0.5) will likely give two same sized groups
  # de.prob - probability that genes will be DE by group(?) i.e. that they'll differ more than just noise
  # nGenes - number of genes to simulate
  # proj_name - character name of dataset or for the parameter object
  # all else we leave as default values. 
  
  # a bit of input validation
  stopifnot(
    is.numeric(batchCells),
    #is.numeric(group.prob),
    is.numeric(de.prob),
    is.numeric(nGenes),
    is.character(proj_name)
  ) 
  
  # initialise and set params
  params <- newSplatParams()
  
  params <- setParams(params, update = list(
    batchCells = batchCells, 
    group.prob = group.prob, 
    de.prob = de.prob,
    nGenes = nGenes
    ))
  
  # simulate based on those parameters
  sim <- splatSimulate(params, method = "groups")
  
  counts_matrix <- counts(sim)
  
  # extract true labelling for later
  true_labels <- data.table(
    cell_id = colnames(counts_matrix),
    true_group = colData(sim)$Group
  )
  
  # this is what we will write out, and then create seurat object as needed for clustering from the counts, and will use true labels for the arandi later.
  return(list(
    counts = counts_matrix,
    labels = true_labels
  ))
  
}



```



```{r Make Parameter Matrix}

parameter_matrix <- tibble(
  ID = 1:10,
  proj_name = paste0("set", sprintf("%02d", 1:10)),
  batchCells = rep(500, 10),
  group.prob = list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ),
  de.prob = c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5),
  nGenes = rep(2000, 10)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/metadata_acs")) {
  
  cat("metadata_acs file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix, file = "D:/AlgorithmComparisonSim/metadata_acs")
  
  cat("metadata file created. \n")
  
}

```

```{r Generate Datasets}

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  # check if there is a file in the directory by the name we want
  
  if (!file.exists(file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i])) | 
      !file.exists(file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i], "true_labels"))) {
    
    # make the dataset if at least one of them is missing
    temp_object <- make_a_dataset(
      proj_name = metadata$proj_name[i],
      batchCells = metadata$batchCells[i],
      group.prob = metadata$group.prob[[i]],
      de.prob = metadata$de.prob[i],
      nGenes = metadata$nGenes[i]
    )
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i]))) {
      cat("File for", i, "already exists, moving on... \n")
    } else {
      
      #write out counts
      fwrite(temp_object$counts, file = paste0("D:/AlgorithmComparisonSim/", metadata[i, 2]))
      
      cat("File: ", metadata$proj_name[i], "created. Moving on... \n")
      
    }
    
    
    
    if (file.exists(file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i], "true_labels"))) {
      cat("File: ", metadata$proj_name[i], "true_labels already exists in folder. Moving on... \n")
      
      } else {
        
      
        fwrite(temp_object$labels, file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i], "true_labels"))  
        
        cat("True labels for", i, "created... \n")
        
    }
    
  }
  
}

```

Since we chose to store the counts instead of the whole seurat object, we will write a quick helper function that brings us to the seurat object. Because of the scaling etc, at this stage we make the decision to cluster all of our methods on the seurat's scaled data layer as far as possible. This will involve some further layers (graphs) for the Louvain and Leiden algorithms but that's fine. 


First order of business - increase amount of bootstrapping, drastically decrease amount of PCs, neglect pure kmeans (without PCA) for now, for timesaving reasons. 

Ok we need to manage to do the clusGap bit. 

```{r More Functions}

# rewrite our K Means function. Need to fully understand what's going on. 

run_kmeans <- function(set_index, counts, k = NULL) {
  
  start_time <- Sys.time()
  cat("BEGIN run_kmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes
  counts <- t(counts)
  
  # we remove dimensions that dont vary at all, as they won't help with clustering
  counts <- counts[, apply(counts, 2, var) > 0]
  
  # I think that we want to scale the data in order for us to see the clusters
  # counts <- scale(counts)
  
  # Estimate k with clusGap
  if (is.null(k)) {
    
    cat("Estimating K... \n")
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = counts,
      FUNcluster = kmeans,
      K.max = 10,
      B = 5, #REDUCED TEMPORARILY, RESET TO 20
      nstart = 5 #REDUCED TEMPORARILY< RESET TO 25
    )
    
    use_k <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
    
    time_2 <- Sys.time()
    cat("took ", time_2 - time_1, " to estimate K as ", use_k, " \n")
    
  } else {
    
    use_k <- k
    cat("K passed as", use_k)
    
  }
  
  # having now estimated or been passed k, we can run kmeans on that same sct_scaled layer. 
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(counts,
                     centers = use_k,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  # and now having produced our cluster assignments we write them out. For now I think I'll only need the cluster assignments, so we literally just write those out. 
  return(kmobject$cluster)
  
  cat("END run_kmeans function. \n")
  
}

####################


# Takes a matrix of counts data (transposes it) and then finds a good number of PCs by which to split the data for cluster evaluation. Returns a list of this number and a plot of the variance explained by each PC, so that we can easily compare whether it's doing the right thing. 
find_npcs <- function(counts) {
  
  counts <- t(counts)
  prcobject <- prcomp(counts)
  variances <- prcobject$sdev^2
  n_pcs <- length(variances)
  
  x <- 1:n_pcs
  y <- variances
  
  x1 <- x[1]
  y1 <- y[1]
  x2 <- x[n_pcs]
  y2 <- y[n_pcs]
  
  a <- y2 - y1
  b <- -(x2 - x1)
  c <- (x2 - x1) * y1 - (y2 - y1) * x1
  
  distances <- abs(a * x + b * y + c) / sqrt(a^2 + b^2)
  
  # The elbow is at the point with maximum distance
  elbow_idx <- which.max(distances)
  
  elbowplot <- ggplot(mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_line() +  # Optional: connect the points
  geom_abline(intercept = y[1], 
              slope = (y[2] - y[1]) / n_pcs,
              linetype = "dashed", 
              color = "gray50") +
  theme_minimal()
  
  return(list(elbow_idx, elbowplot))
  
  cat("END find_npcs function. \n")
  
}

####################


estimate_k <- function(embeddings, scale) {
  
  if (!is.logical(scale) | scale == FALSE) {
    cat("Not scaling embeddings... \n")
  } else if (scale == TRUE) {
    embeddings <- scale(embeddings)
    cat("Scaling embeddings... \n")
  }
    
  clusGapcall <- clusGap(embeddings,
                       FUNcluster = kmeans,
                       nstart = 25,
                       iter.max = 20,
                       K.max = 10,
                       B = 100)

  k_val <- maxSE(clusGapcall$Tab[, "gap"], clusGapcall$Tab[, "SE.sim"], 
             method = "globalmax")

  cat("clusGap gives ", k_val, " as our K estimate. \n")
  
  return(k_val)
  
}



####################


# Takes a seurat object post-SCTransform. Estimates a workable number of PCs, runs PCA, clusters on that number of PCs.

run_PCAkmeans <- function(set_index, counts, k = NULL, scale = FALSE) {
  
  start_time <- Sys.time()
  cat("BEGIN run_PCAkmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes
  counts <- t(counts)
  
  counts <- counts[, apply(counts, 2, var) > 0]

  npcs <- find_npcs(counts)[[1]]
  
  cat("Continuing on ", npcs, " PCs. \n")
  
  embeddings <- prcomp(counts)$x[, 1:npcs]

  if (is.null(k)) {
    
    time_1 <- Sys.time()
    
    use_k <- estimate_k(embeddings = embeddings, scale)
    
    time_2 <- Sys.time()
    cat("Took ", time_2 - time_1, " to estimate k. \n")
    
  } else {
    
    use_k <- k
    cat("K passed as ", use_k, " \n")
    
  }
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(embeddings,
                     centers = use_k,
                     iter.max = 20,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  #r returning just the cluster vector for now
  
  clustering <- data.frame(
    cell = names(kmobject$cluster),
    cluster = kmobject$cluster,
    row.names = NULL
  )
  
  return(clustering)
  
  cat("END run_PCAkmeans function. \n")
  
}


#################


# Takes a seurat object post-SCTransform. Runs the Louvain Algorithm native to Seurat, IF we can make sure it's not doing any extra optimisation under the hood. Otherwise, we can pull the count data, make the SNN graph and cluster ourselves.

run_graph_based <- function(set_index, counts, mode) {
  # set_index - the index of the dataset to be used
  # counts - the counts dataset to be used
  # mode - 1 for Louvain, 2 for Leiden
  
  if (mode == "Louvain") {
    # choose Louvain
    algorithm <- 1
  } else if (mode == "Leiden") {
    # choose Leiden
    algorithm <- 4
  } else {
    algorithm <- 1
    cat("Mode not specified correctly - using Louvain. \n")
  }
  
  start_time <- Sys.time()
  cat("BEGIN run_graph_based function. \n")
  
  # firstly, we want to take our counts and then make it into a seurat object.
  sobject <- CreateSeuratObject(counts = counts, assay = "RNA", project = paste0("set", set_index))
  
  # then, we want to run SCTransform on the data. To be clear, this does normalisation, scaling, and finding variable features. It creates a whole new assay, which we make sure to reference in further steps. 
  
  sctobject <- SCTransform(sobject)
  
  sct_scaled <- LayerData(object = sctobject, assay = "SCT", layer = "scale.data")
  
  # npcs <- find_npcs(sct_scaled)[[1]]
  
  pca_call <- prcomp(sct_scaled, center = TRUE, scale. = FALSE)
  var_explained <- pca_call$sdev^2 / sum(pca_call$sdev^2)
  cumulative_var <- cumsum(var_explained)

  npcs <- which(cumulative_var >= 0.8)[1]

  cat("Continuing Louvain on ", npcs, " PCs... \n")
  
  sctobject <- sctobject |> 
    RunPCA(npcs = npcs) |> 
    FindNeighbors(dims = 1:npcs)
    
  # select Louvain with Algorithm = 1
  sctobject <- FindClusters(sctobject, algorithm = algorithm)
  
  clustering <- data.frame(cell = names(sctobject$seurat_clusters), cluster = as.character(sctobject$seurat_clusters))
  
  cat("END run_louvain function. \n")
  
  return(clustering)
  
  
  
}

```



```{r Louvain Cluster and Writeout}

t_0 <- Sys.time()
set.seed(1618)
cat("BEGINNING LOUVAIN FULL RUN. \n")

for (i in 1:10) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/set", sprintf("%02d", i)))

  temp_cluster <- run_graph_based(i, temp_set, mode = "Louvain")
  
  if (file.exists(
    
    paste0("D:/AlgorithmComparisonSim/LV", sprintf("%02d", i))
    
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/LV", sprintf("%02d", i)), "already exists... \n") 
  } else {
      
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/LV", sprintf("%02d", i)))
    
    cat("Written out file from set", i, " ... \n")
    
    }

}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 10: ", t_1 - t_0)

```

```{r Leiden Cluster and Writeout}

t_0 <- Sys.time()
set.seed(1618)
cat("BEGINNING LEIDEN FULL RUN. \n")

for (i in 1:10) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/set", sprintf("%02d", i)))

  temp_cluster <- run_graph_based(i, temp_set, mode = "Leiden")
  
  if (file.exists(
    
    paste0("D:/AlgorithmComparisonSim/LD", sprintf("%02d", i))
    
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/LD", sprintf("%02d", i)), "already exists... \n") 
  } else {
      
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/LD", sprintf("%02d", i)))
    
    cat("Written out file from set", i, " ... \n")
    
    }

}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 10: ", t_1 - t_0)

```

```{r KMeans Cluster and Writeout}

t_0 <- Sys.time()

for (i in 1:10) {
  
  temp_set <- fread(paste0(
    "D:/AlgorithmComparisonSim/set", sprintf("%02d", i)
  ))
  
  #trying WITHOUT scaling first
  temp_cluster <- run_PCAkmeans(i, temp_set, scale = FALSE)
  
  if (file.exists(
    
    paste0("D:/AlgorithmComparisonSim/PK", sprintf("%02d", i))
    
  )) { 
    cat("File ", paste0("D:/AlgorithmComparisonSim/PK", sprintf("%02d", i)), "already exists... \n") 
  } else {
      
    fwrite(temp_cluster, paste0("D:/AlgorithmComparisonSim/PK", sprintf("%02d", i)))
    
    cat("Written out file from set", i, " ... \n")
    
    }
  
  cat("FINISHED clustering the ", i, "th one! \n")
    
}

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do that. \n")

```

Creating clustering evaluation function. Choose Silhouette, Davies-Bouldin, Calinski-Harabasz, Adj Rand Index. 

```{r Cluster Evaluation Function}

cluster_scores <- function(clustering, original_counts, true_labels) {
  
  # silhouette score
  whole_silhouette <- silhouette(clustering, dist = dist(original_counts))
  score_silhouette <- mean(whole_silhouette[, 3])
  
  #calinhara 
  score_calinhara <- calinhara(original_counts, clustering)
  
  #dbouldin
  score_dbouldin <- index.DB(original_counts, clustering)
  
  #arandi
  score_ari <- mclust::adjustedRandIndex(true_labels, clustering)
  
  
  
}

```




