---
title: "AlgorithmComparisonSim"
output: html_document
---

```{r Libraries}

library(tidyverse)
library(scater)
library(splatter)
library(Seurat)
library(cluster)
library(data.table)

```

```{r Functions}


# takes parameters specified and returns a number of seurat objects to those specifications, just in a list. they'll be slightly different since theres randomness and we make sure the random seed is different each time. 

make_a_dataset <- function(proj_name, batchCells, group.prob, de.prob, nGenes) {
  
  # batchCells - a vector giving the number of cells per batch, ie two batches with 100 cells is (100, 100)
  # group.prob - vector specifying probabilities per group, ie (0.5, 0.5) will likely give two same sized groups
  # de.prob - probability that genes will be DE by group(?) i.e. that they'll differ more than just noise
  # nGenes - number of genes to simulate
  # proj_name - character name of dataset or for the parameter object
  # all else we leave as default values. 
  
  # a bit of input validation
  stopifnot(
    is.numeric(batchCells),
    #is.numeric(group.prob),
    is.numeric(de.prob),
    is.numeric(nGenes),
    is.character(proj_name)
  ) 
  
  # initialise and set params
  params <- newSplatParams()
  
  params <- setParams(params, update = list(
    batchCells = batchCells, 
    group.prob = group.prob, 
    de.prob = de.prob,
    nGenes = nGenes
    ))
  
  # simulate based on those parameters
  sim <- splatSimulate(params, method = "groups")
  
  counts_matrix <- counts(sim)
  
  # this is what we will write out, and then create seurat object as needed for clustering
  return(counts_matrix)
  
}



```



```{r Make Parameter Matrix}

parameter_matrix <- tibble(
  ID = 1:10,
  proj_name = paste0("set", sprintf("%02d", 1:10)),
  batchCells = rep(500, 10),
  group.prob = list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ),
  de.prob = c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5),
  nGenes = rep(2000, 10)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/metadata_acs")) {
  
  cat("metadata_acs file already exists. \n")
  
} else {
  
  fwrite(parameter_matrix, file = "D:/AlgorithmComparisonSim/metadata_acs")
  
  cat("metadata file created. \n")
  
}

```

```{r Generate Datasets}

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    nGenes = as.numeric(nGenes),
    group.prob = strsplit(group.prob, "\\|") |>
      lapply(as.numeric)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  # check if there is a file in the directory by the name we want
  
  if (file.exists(
    
    file = paste0("D:/AlgorithmComparisonSim/", metadata$proj_name[i])
    
  )) {
    
    cat("File: ", metadata$proj_name[i], "already exists in folder. Moving on... \n")
    
  } else {
    
    temp_object <- make_a_dataset(
      proj_name = metadata$proj_name[i],
      batchCells = metadata$batchCells[i],
      group.prob = metadata$group.prob[[i]],
      de.prob = metadata$de.prob[i],
      nGenes = metadata$nGenes[i]
    )
    
    cat("This df has", ncol(temp_object), "cols and ", nrow(temp_object), "rows. \n")
    
    fwrite(temp_object, file = paste0("D:/AlgorithmComparisonSim/", metadata[i, 2]))
    
    cat("File: ", metadata$proj_name[i], "created. Moving on... \n")
    
  }
  
}

```

Since we chose to store the counts instead of the whole seurat object, we will write a quick helper function that brings us to the seurat object. Because of the scaling etc, at this stage we make the decision to cluster all of our methods on the seurat's scaled data layer as far as possible. This will involve some further layers (graphs) for the Louvain and Leiden algorithms but that's fine. 


First order of business - increase amount of bootstrapping, drastically decrease amount of PCs, neglect pure kmeans (without PCA) for now, for timesaving reasons. 

Ok we need to manage to do the clusGap bit. 

```{r More Functions}

# rewrite our K Means function. Need to fully understand what's going on. 

run_kmeans <- function(set_index, counts, k = NULL) {
  
  start_time <- Sys.time()
  cat("BEGIN run_kmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes
  counts <- t(counts)
  
  # we remove dimensions that dont vary at all, as they won't help with clustering
  counts <- counts[, apply(counts, 2, var) > 0]
  
  # I think that we want to scale the data in order for us to see the clusters
  # counts <- scale(counts)
  
  # Estimate k with clusGap
  if (is.null(k)) {
    
    cat("Estimating K... \n")
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = counts,
      FUNcluster = kmeans,
      K.max = 10,
      B = 5, #REDUCED TEMPORARILY, RESET TO 20
      nstart = 5 #REDUCED TEMPORARILY< RESET TO 25
    )
    
    use_k <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
    
    time_2 <- Sys.time()
    cat("took ", time_2 - time_1, " to estimate K as ", use_k, " \n")
    
  } else {
    
    use_k <- k
    cat("K passed as", use_k)
    
  }
  
  # having now estimated or been passed k, we can run kmeans on that same sct_scaled layer. 
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(counts,
                     centers = use_k,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  # and now having produced our cluster assignments we write them out. For now I think I'll only need the cluster assignments, so we literally just write those out. 
  return(kmobject$cluster)
  
  cat("END run_kmeans function. \n")
  
}

####################


# Takes a matrix of counts data (transposes it) and then finds a good number of PCs by which to split the data for cluster evaluation. Returns a list of this number and a plot of the variance explained by each PC, so that we can easily compare whether it's doing the right thing. 
find_npcs <- function(counts) {
  
  counts <- t(counts)
  prcobject <- prcomp(counts)
  variances <- prcobject$sdev^2
  n_pcs <- length(variances)
  
  x <- 1:n_pcs
  y <- variances
  
  x1 <- x[1]
  y1 <- y[1]
  x2 <- x[n_pcs]
  y2 <- y[n_pcs]
  
  a <- y2 - y1
  b <- -(x2 - x1)
  c <- (x2 - x1) * y1 - (y2 - y1) * x1
  
  distances <- abs(a * x + b * y + c) / sqrt(a^2 + b^2)
  
  # The elbow is at the point with maximum distance
  elbow_idx <- which.max(distances)
  
  elbowplot <- ggplot(mapping = aes(x = x, y = y)) +
  geom_point() +
  geom_line() +  # Optional: connect the points
  geom_abline(intercept = y[1], 
              slope = (y[2] - y[1]) / n_pcs,
              linetype = "dashed", 
              color = "gray50") +
  theme_minimal()
  
  return(list(elbow_idx, elbowplot))
  
  cat("END find_npcs function. \n")
  
}



####################


# Takes a seurat object post-SCTransform. Estimates a workable number of PCs, runs PCA, clusters on that number of PCs.

run_PCAkmeans <- function(set_index, counts, k = NULL) {
  
  start_time <- Sys.time()
  cat("BEGIN run_PCAkmeans function. \n")
  
  # We want to transpose the matrix so it's cells x genes
  counts <- t(counts)
  
  counts <- counts[, apply(counts, 2, var) > 0]

  npcs <- find_npcs(counts)[[1]]
  
  cat("Continuing on ", npcs, " PCs. \n")
  
  embeddings <- prcomp(counts)$x[, 1:npcs]
  
  # run kmeans

  if (is.null(k)) {
    
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = embeddings,
      FUNcluster = kmeans,
      K.max = 10,
      B = 200, 
      nstart = 25
    )
    
    use_k <- maxSE(clusGapcall$tab[, "gap"],
                   clusGapcall$tab[, "SE.sim"],
                   method = "Tibs2001SEmax")
    
    time_2 <- Sys.time()
    cat("Took ", time_2 - time_1, " to estimate k.")
    
  } else {
    
    use_k <- k
    cat("K passed as ", use_k, " \n")
    
  }
  
  time_3 <- Sys.time()
  cat("Beginning KMeans clustering on ", use_k, "centres... \n")
  
  kmobject <- kmeans(embeddings,
                     centers = use_k,
                     nstart = 25)
  
  time_4 <- Sys.time()
  cat("Took ", time_4 - time_3, "to finish clustering... \n")
  
  #r returning just the cluster vector for now
  return(kmobject$cluster)
  
  cat("END run_PCAkmeans function. \n")
  
}


#################


# Takes a seurat object post-SCTransform. Runs the Louvain Algorithm native to Seurat, IF we can make sure it's not doing any extra optimisation under the hood. Otherwise, we can pull the count data, make the SNN graph and cluster ourselves.

run_louvain <- function(set_index, counts, resolution = NULL) {
  
  start_time <- Sys.time()
  cat("BEGIN run_louvain function. \n")
  
  # firstly, we want to take our counts and then make it into a seurat object.
  sobject <- CreateSeuratObject(counts = counts, assay = "RNA", project = paste0("set", set_index))
  
  # then, we want to run SCTransform on the data. To be clear, this does normalisation, scaling, and finding variable features. It creates a whole new assay, which we make sure to reference in further steps. 
  
  sctobject <- SCTransform(sobject)
  
  sct_scaled <- LayerData(object = sctobject, assay = "SCT", layer = "scale.data")
  
  pca_call <- prcomp(sct_scaled, center = TRUE, scale. = TRUE)
  var_explained <- pca_call$sdev^2 / sum(pca_call$sdev^2)
  cumulative_var <- cumsum(var_explained)
  
  npcs <- which(cumulative_var >= 0.8)[1]

  cat("Continuing Louvain on ", npcs, " PCs... \n")
  
  sctobject <- sctobject |> 
    RunPCA(npcs = npcs) |> 
    FindNeighbors(dims = 1:npcs)
    
  # select Louvain with Algorithm = 1
  sctobject <- FindClusters(sctobject, algorithm = 1)
  
  clustering <- sctobject$seurat_clusters
  
  return(clustering)
  
  cat("END run_louvain function. \n")
  
}

```


So far we're seeing the whole thing does run, but the clusgap is returning one centre. am going to run it through all the sets to see if it's always doing this, and then try and tweak it. it will unfortunately probably take 5ish minutes per one, so we'll go with just the first five or so.

Yep so they are all on one centre. So we need to check what's going wrong with our clusgap call, or maybe it's just like too many dimensions. who knows eh.

Attempting with the true k passed gives us that kmeans is terrible on the whole data and on 20 PCS it's pretty mid. Now we go to louvain and leiden, which do not suffer from this need to estimate k, and will probably show that these outperform KMeans and KMeans + PCA. To note is that for Louvain and Leiden, they are performed on a PCA space by default. I.e., there is a silent ( + PCA) on each one. 

```{r KMeans Cluster and Writeout}

t_0 <- Sys.time()

set05 <- fread("D:/AlgorithmComparisonSim/set05")

kmeans01 <- run_PCAkmeans(set_index = 5, counts = set05)

t_1 <- Sys.time()
cat("Took ", t_1 - t_0, "to do that. \n")

```

```{r Louvain Cluster and Writeout}

t_0 <- Sys.time()

for (i in 1:10) {
  
  temp_set <- fread(paste0("D:/AlgorithmComparisonSim/set", sprintf("%02d", i)))

  temp_cluster <- run_louvain(i, temp_set)
  
  if (file.exists(
    
    paste0()
    
  ))

}

t_1 <- Sys.time()
cat("TOTAL TIME FOR ALL 5: ", t_1 - t_0)

```




