---
title: "AlgorithmComparisonSim"
output: html_document
---

```{r Libraries}

library(tidyverse)
library(scater)
library(splatter)
library(Seurat)
library(cluster)

```

```{r Functions}


# takes parameters specified and returns a number of seurat objects to those specifications, just in a list. they'll be slightly different since theres randomness and we make sure the random seed is different each time. 

make_a_dataset <- function(proj_name, batchCells, group.prob, de.prob, nGenes) {
  
  # batchCells - a vector giving the number of cells per batch, ie two batches with 100 cells is (100, 100)
  # group.prob - vector specifying probabilities per group, ie (0.5, 0.5) will likely give two same sized groups
  # de.prob - probability that genes will be DE by group(?) i.e. that they'll differ more than just noise
  # nGenes - number of genes to simulate
  # proj_name - character name of dataset or for the parameter object
  # all else we leave as default values. 
  
  # a bit of input validation
  stopifnot(
    is.numeric(batchCells),
    #is.numeric(group.prob),
    is.numeric(de.prob),
    is.numeric(nGenes),
    is.character(proj_name)
  ) 
  
  # initialise and set params
  params <- newSplatParams()
  
  params <- setParams(params, update = list(
    batchCells = batchCells, 
    group.prob = group.prob, 
    de.prob = de.prob,
    nGenes = nGenes
    ))
  
  # simulate based on those parameters
  sim <- splatSimulate(params, method = "groups")
  
  counts_matrix <- counts(sim)
  
  # this is what we will write out, and then create seurat object as needed for clustering
  return(counts_matrix)
  
}



```

At present my biggest concern with the below data is that the SCTransform is hanging on to only the 100 most variable genes. I have not found this answer immediate from documentation so I need to dig a bit more for that, or maybe am missing something. 

the columns i want to have in param matrix are:
ID, projname, batchcells, group prob, de prob, nGenes (proj name will be the same as filename)

```{r Make Parameter Matrix}

parameter_matrix <- tibble(
  ID = 1:10,
  proj_name = paste0("set", 1:10),
  batchCells = rep(500, 10),
  group.prob = list(
    c(0.25, 0.25, 0.25, 0.25),
    c(0.33, 0.33, 0.34),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.3, 0.3, 0.4),
    c(0.25, 0.25, 0.25, 0.25),
    c(0.2, 0.3, 0.5),
    c(0.15, 0.15, 0.2, 0.25, 0.25),
    c(0.4, 0.3, 0.3),
    c(0.2, 0.2, 0.2, 0.2, 0.2),
    c(0.25, 0.35, 0.4)
  ),
  de.prob = c(0.05, 0.1, 0.1, 0.2, 0.2, 0.2, 0.2, 0.3, 0.3, 0.5),
  nGenes = rep(2000, 10)
)

# write out this file
if (file.exists(file = "D:/AlgorithmComparisonSim/metadata_acs")) {
  
  cat("metadata_acs file already exists")
  
} else {
  
  fwrite(parameter_matrix, file = "D:/AlgorithmComparisonSim/metadata_acs")
  
}

```

```{r Generate Datasets}

# read the file back in and set random seed for reproducibility
metadata <- fread("D:/AlgorithmComparisonSim/metadata_acs")

metadata <- metadata |> 
  mutate(
    ID = as.numeric(ID),
    proj_name = as.character(proj_name),
    batchCells = as.numeric(batchCells),
    de.prob = as.numeric(de.prob),
    group.prob = lapply(group.prob, as.numeric),
    nGenes = as.numeric(nGenes)
  )

set.seed(1618)

for (i in 1:nrow(metadata)) {
  
  # check if there is a file in the directory by the name we want
  
  if (file.exists(
    
    file = paste0("D:/AlgorithmComparisonSim/", metadata[i, 2])
    
  )) {
    
    cat("File: ", metadata[i, 2], "already exists in folder. Moving on... \n")
    
  } else {
    
    temp_object <- do.call(make_a_dataset, metadata[i, 2:6])
    
    fwrite(temp_object, file = paste0("D:/AlgorithmComparisonSim/", metadata[i, 2]))
    
    cat("File: ", metadata[i, 2], "created. Moving on... \n")
    
  }
  
}

```

```{r More Functions}

# take SCTransformed seurat (i.e ready for dim red and clustering) and then run KMeans on its counts matrix with no dimensionality reduction

run_kmeans <- function(sct_seurat_object, k = NULL) {
  
  start_time <- Sys.time()
  
  sct_scaled <- LayerData(
    object = sct_seurat_object, 
    assay = "SCT", 
    layer = "scale.data")

  if (is.numeric(k) == TRUE) {
    centers = k
  } else {
    
    time_1 <- Sys.time()
    
    clusGapcall <- clusGap(
      x = sct_scaled,
      FUNcluster = kmeans,
      K.max = 10,
      B = 20
    )
    
    centers <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
    
    time_2 <- Sys.time()
    cat("took ", time_2 - time_1, " to estimate k \n")
    
  }
  
  time_3 <- Sys.time()
  
  # run kmeans:
  kmobject <- kmeans(sct_scaled,
                     centers = centers,
                     nstart = 25)
  
  end_time <- Sys.time()
  cat("took ", end_time - time_3, " to cluster and ", end_time - start_time, "in total \n")
  
  return(kmobject)
  
}

# Takes a seurat object post-SCTransform. Estimates a workable number of PCs, runs PCA, clusters on that number of PCs.

run_PCAkmeans <- function(sct_seurat_object, k = NULL) {
  
  # sct_seurat_object - object of class seurat with SCTransform applied
  # k - centers for KMeans. If left NULL, estimated with Gap Statistic
  
  PCA_sct_seurat_object <- RunPCA(sct_seurat_object, npcs = 100, verbose = TRUE)
  
  # pull PCA embeddings
  embeddings <- Embeddings(PCA_sct_seurat_object)
  
  # run kmeans

  if (is.numeric(k) == TRUE) {
    centers = k
  } else {
    clusGapcall <- clusGap(
      x = embeddings,
      FUNcluster = kmeans,
      K.max = 10,
      B = 5
    )
    
    centers <- maxSE(clusGapcall$Tab[, "gap"],
                     clusGapcall$Tab[, "SE.sim"],
                     method = "firstSEmax")
  }
  
  # run kmeans:
  kmobject <- kmeans(embeddings,
                     centers = centers,
                     nstart = 25)
  
  return(kmobject)
  
}

# Takes a seurat object post-SCTransform. Runs the Louvain Algorithm native to Seurat, IF we can make sure it's not doing any extra optimisation under the hood. Otherwise, we can pull the count data, make the SNN graph and cluster ourselves.

run_louvain <- function(sct_seurat_object, resolution  = NULL) {
  
  sct_seurat_object <- sct_seurat_object |> 
    RunPCA(npcs = 100) |> 
    FindNeighbors(dims = 1:30)
    
  
  
  lv_sct <- FindClusters(sct_seurat_object, algorithm = 1)
  
  return(lv_sct)
  
}

```

```{r Simulate}

kmeans_sct_sim1 <- run_kmeans(sct_sim1)

pca_kmeans_sct_sim1 <- run_PCAkmeans(sct_sim1)

lv_sct_sim1 <- run_louvain(sct_sim1)


# for the louvain clustering

p_sct_sim1 <- sct_sim1 |> 
    RunPCA(npcs = 100) |> 
    FindNeighbors(dims = 1:30)

embeddings <- Embeddings(p_sct_sim1, reduction = "pca")[, 1:30]

lv_clusters <- as.numeric(Idents(sct_sim1))
dist_mat <- dist(embeddings)
lv_sil <- silhouette(lv_clusters, dist_mat)
lv_avg_sil <- mean(lv_sil[, 3])

# then need to run louvain on this




```

```{r Write}



```



