---
title: "KMeansCurseSim"
output: html_document
---

Here, we're going to copy across the relevant parts of "takefour.Rmd". Then, we will write a workflow in pseudocode, neaten them up, and produce a graph with one dataset (artificial, sliced), one algorithm (kmeans), and one metric (Adjusted Rand Index). Once we've done this and sufficiently observed the curse of dimensionality with it, IF it's quick and easy, we can move on to adding to either of those three sections. 

It will be REALLY nice to have this bit done such that we can write about it. Ok. We begin. 

```{r Libraries}

library(tidyverse) #for general stuff
library(factoextra) #for clustering
library(cluster) #for clustering
library(mclust) #for more clustering specifically the arandi

```

```{r Functions}

# m - number of rows per cluster
# n - number of dimensions for the dataset (number of columns)
# N - noise dimensions to be added to the dataset 'in post' of the generation
# k - clusters to generate
# Overlap - boolean variable, whether the variances of each cluster should be inflated s.t. they overlap, or should they be well-separated?
# Equal-In - boolean - should my clusters be spherical, or 'ovoid'?
simple_data <- function(m, n, N, k, overlap, equalin) {
  
  # make sure nothing's gonna break
  stopifnot(
    is.logical(overlap), 
    is.logical(equalin)
    #is.logical(equalout)
  )
  
  # check some options based on our arguments
  if (overlap == TRUE) {
    p_var <- rep(0.5, k) 
  } else {
    p_var <- rep(0.2, k)
  }
  
  if (equalin == FALSE) {
    p_var <- abs(matrix(rnorm(k, p_var[[1]], 0.25),  nrow = 1, ncol = k)) # might need slight adjusting
    
  }
  
  #if (equalout)
  
  unmerged <- list()
  
  # create centre data as separate matrices in 'unmerged'
  for (i in 1:k) {
    cat("generating cluster ", i, "... \n")
    unmerged[[i]] <- matrix(rnorm(m*n, i, p_var[[i]]), nrow = m, ncol = n)
    
    noise <- matrix(rnorm(N*m, mean = 0, sd = 0.3), nrow = m, ncol = N)
    unmerged[[i]] <- cbind(unmerged[[i]], noise)
    
    unmerged[[i]] <- cbind(unmerged[[i]], matrix(data = as.numeric(i), nrow = m, ncol = 1)) # add tag column
    
    cat("finished cluster ", i, "... \n")
  }
  
  cat("finished all clusters... \n")
  
  frame <- as.data.frame(do.call(rbind, unmerged)) # smush the matrices together
  
  names(frame)[1:n] <- paste0("dim", 1:n)
  if (N != 0) {
    names(frame)[(n + 1):(n + N)] <- paste0("noise", 1:N)
  }
  
  names(frame)[ncol(frame)] <- "Tag"
  
  return(frame)
  
}

# uses simple_data to slice, cluster and score a pssed dataset.
slice_cluster_score <- function(data, v) {
  # data - dataset to be sliced n scored
  # v - vector of dimension numbers to take arandi score at
  
  #get true labels
  true_labels <- data$Tag
  
  #drop the tag col 
  data_no_tag <- select(data, -Tag)
  
  # initialise scores vector
  scores <- numeric(length(v))
  
  # slice the dataset, estimate k with gap statistic, then add the arandi to scores vector
  for (i in seq_along(v)) {
    
    # slice:
    current_slice <- select(data_no_tag, 1:i)
    
    # cluster:
    
    # estimate k:
    clusGapcall <- clusGap(
      x = current_slice,
      FUNcluster = kmeans,
      K.max = 10
    )
    
    k_estimated <- maxSE(clusGapcall$Tab[, "gap"],
                         clusGapcall$Tab[, "SE.sim"], 
                         method = "firstSEmax")
    
    # run kmeans:
    kmobject <- kmeans(current_slice, 
                       k_estimated,
                       nstart = 25)
    
    scores[i] <- mclust::adjustedRandIndex(true_labels, kmobject$cluster)
    
    #cat("Dimension: ", v[i], "gives arandi score: ", round(scores[i], 4), " \n")
    
  }
  
  # format data frame for output
  result <- data.frame(
    dimension = v,
    score = scores
  )
  
  return(result)
  
}



```

```{r Making the Data for the Plot}

log_sequence <- unique(round(10^seq(1, 4, length.out = 30)))

test_data <- simple_data(200, 10, 10^4 - 10, 4, overlap = FALSE, equalin = TRUE)

trial <- slice_cluster_score(test_data, log_sequence)

```


```{r Plot, echo=TRUE}

plot1 <- ggplot(data = trial, mapping  = aes(dimension, score)) +
  geom_point() +
  geom_smooth() +
  scale_x_log10() +
  xlab("Dimension") +
  ylab("Adjusted Rand Index") +
  labs(title = "Adjusted Rand Index vs Dimension")

plot1

```

