---
title: "KMeansCurseSim"
output: html_document
---

Here, we're going to copy across the relevant parts of "takefour.Rmd". Then, we will write a workflow in pseudocode, neaten them up, and produce a graph with one dataset (artificial, sliced), one algorithm (kmeans), and one metric (calinhara). Once we've done this and sufficiently observed the curse of dimensionality with it, IF it's quick and easy, we can move on to adding to either of those three sections. 

It will be REALLY nice to have this bit done such that we can write about it. Ok. We begin. 

```{r Libraries}

library(tidyverse) #for general stuff
library(factoextra) #for clustering
library(cluster) #for clustering
library(mclust) #for more clustering specifically the arandi

```

```{r Functions}

# Creates a centre of dimension n, centred at the n-dim point where each entry 
# is x, and adds N dimensions of gaussian noise
sample_maker_noisy <- function(n, m, iteration, N) {
  # n - useful features
  # m - observations per feature
  # x - symmetrical location of the centre
  # N - number of noise dimensions
  
  temp_mean <- iteration*1.5
  temp_sd <- 2
  
  # create the useful and noisy dimensions separately then combine into one matrix
  mat_base <- matrix(rnorm(n*m, temp_mean, temp_sd), nrow = m, ncol = n)
  mat_noise <- matrix(rnorm(N*m, 0, 2), nrow = m, ncol = N)
  mat <- cbind(mat_base, mat_noise)
  
  # format data frame for output - add tag variable s.t. we can compute arandi later
  df <- as.data.frame(mat)
  names(df) <- paste0("var", seq_len(n))
  df$tag <- iteration
  
  return(df)
}


# uses sample_maker_noisy to create one dataframe of k 'noisy' centres
create_data_noisy <- function(k, n, m, N) {
  # k - number of centres
  # n - number of meaningful features 
  # m - observations per feature
  # N - number of features of noise (so the data will have n + N features in total)
  # we expect N >> n
  
  frame_list <- list()
  
  # iterate through k centres and create the data for each
  for (i in 1:k) {
    frame_list[[i]] <- sample_maker_noisy(n = n, m = m, iteration = i, N = N)
    print(paste0("MADE DATASET", i, "OF", k))
  }
  
  #make into one dataframe again with do.call
  df <- data.frame(do.call(rbind, frame_list))
  
  return(df)
}

# uses create_data_noisy to generate adjusted rand index scores per a passed vector of dimensions
scores_generator_arandi_noisy <- function(k, n, m, N, v) {
  # k - number of centres
  # n - number of meaningful dimensions
  # m - observations per centre
  # N - number of noise dimensions
  # v - vector of dimension numbers to take arandi score at
  
  #create big dataset
  max_dim = max(v)
  cat("Generating data of", max_dim, "dimensions... \n")
  data <- create_data_noisy(k, n, m, N)
  cat("Completed data generation... \n")
  
  #get true labelling
  true_labels <- data$tag
  
  #haven't done this in the other frames but just in case, drop the tag col 
  data_no_tag <- data[, -which(names(data) == "tag"), drop = FALSE]
  
  scores <- numeric(length(v))
  
  # slice the dataset, kmeans it with true k, then add the arandi to scores vector
  for (i in seq_along(v)) {
    
    current_slice <- data_no_tag[, 1:v[i], drop = FALSE]
    kmobject <- kmeans(current_slice, centers = k, nstart = 10, iter.max = 25)
    
    scores[i] <- mclust::adjustedRandIndex(true_labels, kmobject$cluster)
    
    cat("Dimension: ", v[i], "gives arandi score: ", round(scores[i], 4), " \n")
    
  }
  
  # format data frame for output
  result <- data.frame(
    dimension = v,
    score = scores
  )
  
  return(result)
  
}

```

```{r Making the Data for the Plot}

log_sequence <- unique(round(10^seq(1, 4, length.out = 30)))

output <- scores_generator_arandi_noisy(k = 3, n = 10, m = 200, v = log_sequence, N = max(log_sequence))
```


```{r Plot, echo=TRUE}

plot1 <- ggplot(data = output, mapping  = aes(dimension, score)) +
  geom_point() +
  geom_smooth() +
  scale_x_log10() +
  xlab("Dimension") +
  ylab("Adjusted Rand Index") +
  labs(title = "Adjusted Rand Index vs Dimension")

plot1

```

